Reassessing the Role of Theory and Machine Learning in Forecasting Civil Conflict

### 2020-09-22 paper.Rmd
expand discussion; done for now

Conversely, not all factors that are "theoretically" important are also important for prediction, and some factors that turn out to be good predictors do not originate from prior theory. For example, both @wgb:2007 and @hill:jones:2014 examine a variety of variables that are considered "theoretically" important for explaining civil war onset and state repression, respectively, and find that there is substantial variation in how useful they are for for actually predicting those outcomes.

Another way in which we could contrast theoretical from atheoretical prediction models is in terms of their flexibility in regard to functional form. Theory should not only tell us what factors matter, but also how they relate to each other. We think that few would be strongly committed to the idea that the linear additive equations that dominate published regression analyses, maybe with the occasional polynomial or multiplicative interaction terms, are *the* correct functional specification. Is the danger posed by low-level violence the same when tensions are high, with high demands and high non-violent repression, as when tensions are low? Are those relationships linear in the log-odds? Some machine learning models, random forests included, are flexible in this regard and to some extent capable of learning non-linear and interactive relationships. B&S implicitly concede on this point since they use random forests for both their Escalation model and the atheoretical alternatives. But in the absence of strong a priori theoretical expectations about functional form, why not use machine learning tools to try to uncover any such non-linear relationships in need of explanation, and use this to improve our theory?

>>>>

Furthermore, if what we mean with "theory" is such that it suggests a large range of variables that could be related to an outcome, what do we really gain from having this theory in our predictive model? We can construct a model designed to predict well while avoiding overfitting, and then go back and try to couch it in theoretical language---which, given the malleability of extant theory, we probably could in many cases---but are we then better off? Certainly that's not what B&S are trying to suggest since this would be a problematic way to build a theory. But when theory is extensive or malleable so that it suggests a broad range of variables and specifications, and thus a broad range of specifications could post fact be tied to a theory, it is hard to demonstrate that this theory is in fact not the result of predictive modeling's equivalent to p-hacking, namely developing a model specification and any associated "theory" with any eye towards out-of-sample predictive accuracy [cf. mahmood colarsei paper on how to do this properly though]. Incidentally, this is also a recipe for creating overfit models whose accuracy does not generalize, and whose underlying theories are by extension not externally valid [zones farris paper].

Conversely, not all factors that are "theoretically" important or suggested by the literature are also important for prediction, and some factors that turn out to be good predictors do not originate from prior theory. For example, both @wgb:2007 and @hill:jones:2014 examine a variety of variables that are considered "theoretically" important for explaining civil war onset and state repression, respectively, and find that there is substantial variation in how useful they are for for actually predicting those outcomes. And it is entirely possible that a focus on building accurate predictive models could uncover factors overlooked in the literature, e.g. the strong association between infant mortality and a broad variety of conflict outcomes like political instability [@goldstone:etal:2010], irregular leadership changes [@beger:etal:2017], and coups.^[Andreas Beger and Michael D. Ward, 2020, “Coup forecasts for 2020”, https://www.predictiveheuristics.com/forecasts]

% MIKE: is it actually true that infant mortality has mostly not been studied in the context of explaining conflict risk? Goldstone in their article claim that it has, but for me that's news.

Aside from selection of the variables that go into a model, another way in which we could contrast theoretical from atheoretical prediction models is in terms of their flexibility in regard to functional form. Theory should not only tell us what factors matter, but also how they relate to each other. We think that few would be strongly committed to the idea that the linear additive equations that dominate published regression analyses, maybe with the occasional polynomial or multiplicative interaction terms, are *the* correct functional specification. Is the danger posed by low-level violence the same when tensions are high, with high demands and high non-violent repression, as when tensions are low? Are those relationships linear in the log-odds? Some machine learning models, random forests included, are flexible in this regard and to some extent capable of learning non-linear and interactive relationships. B&S implicitly concede on this point since they use random forests for both their Escalation model and the atheoretical alternatives. But in the absence of strong a priori theoretical expectations about functional form, why not use machine learning tools to try to uncover any such non-linear relationships in need of explanation, and use this to improve our theory?
